---
title: "NHS App Utilisation"
author: "<a href='https://github.com/ASW-Analyst/'>Andy Wilson</a><br><a href='https://transformationunit.nhs.uk/'>NHS Transformation Unit</a>"
date: "08/12/2025"
date-format: "DD MMMM YYYY"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    self-contained: true
    css: ../config/nhs_tu_theme.css
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = '100%'
)

library(here)
source(paste0(here("src", "requirements", "packages.R")))
source(paste0(here("src", "config", "palette.R")))
source(paste0(here("src", "load", "load.R")))
source(paste0(here("src", "processing", "populations_processing.R")))
source(paste0(here("src", "processing", "outpatient_proxy.R")))
source(paste0(here("src", "processing", "mi_processing.R")))
source(paste0(here("src", "processing", "demographics.R")))
source(paste0(here("src", "analysis", "mi.R")))
source(paste0(here("src", "analysis", "demographics.R")))


```


# Introduction
***

The [10-Year Health Plan for England](https://assets.publishing.service.gov.uk/media/6888a0b1a11f859994409147/fit-for-the-future-10-year-health-plan-for-england.pdf) has the ambitions for using NHS App to collect feedback from patients about their care. By 2028 the App will:

> _allow patients to leave feedback on the care they have received - compiled and communicated back to providers, clinical teams and professionals in easy-to-action formats_ 


There is [significant evidence](https://www.kingsfund.org.uk/insight-and-analysis/long-reads/tackling-health-inequalities-seven-priorities-nhs) on the health inequalities that exist in access to care, patient outcomes and experiences. Listening and acting on patient feedback and understanding how services don't meet certain groups needs is critical to addressing these inequalities. Therefore, it is important that feedback collected through the NHS App is representative of the patients who use those services.


# What do we already know about usage of the App?
***
For the last few months NHS England has been publishing some high-level statistics on the number of logins and the number of unique logins to the NHS App. This data is publicly available and the latest publication can be accessed [here](https://digital.nhs.uk/data-and-information/publications/statistical/nhs-app-statistics). These publications highlight that for the latest month of data (`r format(mi_latest[[1, 1]], "%B %Y")`) there were **`r scales::comma(mi_latest[[1,2]])`** logins to the App. Of these, **`r scales::comma(mi_latest[[1,3]])`** were unique users.

The lowest level of granularity the data is published at is for individual Integrated Care Boards (ICBs). Calculating the rate of unique logins for each ICB shows there is significant variation in App utilisation across the country. The ICB with the highest rate, _`r mi_latest_comp[[1, 3]]`_, is **`r round(mi_latest_comp[[1, 11]] / mi_latest_comp[[2, 11]], 1)`** that of the lowest, _`r mi_latest_comp[[2, 3]]`_.

```{r mi_icb_rates_chart}

mi_icb_rates_latest_chart

```



However, there is no publicly available information to help explain the differences between these rates. What are some of the population demographic factors that could be impacting use of the NHS App? How can we be confident that those using the NHS App are representative of the wider population or those who are using NHS services? Without understanding this how can we be confident that feedback on services will reflect the views and needs to those using them?

<br/>


# Demographics FOI Request
***

## FOI Request

### Our Request
We submitted a Freedom of Information Request (FOI) to NHS England requesting the latest utilisation data for the App broken down by the following population demographics:

* __Age__: Age of the patient using the NHS App.
* __Gender__: Breakdown of the gender of patients using the NHS App.
* __Ethnicity__: Data on the ethnicity group of patients using the NHS App, broken down using the groupings within the NHS Data Dictionary or another appropriate grouping.
* __Deprivation__: The Index of Multiple Deprivation (IMD) decile for the user based on their resident postcode. If this wasn't possible to use the IMD decile based on the GP Practice location.

### Data Provided and Caveats
NHS England were able to provide a breakdown of data for March, April and May 2025 for the following:

* __Age__: A breakdown of age into 5-year age bands. For the age band analysis, an initial band of age ‘13 to 14' is presented, since users must be aged 13 or over to register with the NHS App. Users recorded as being aged 12 or younger have been included in the Unknown category.
* __Deprivation__: A breakdown by IMD deciles based on the user's GP practice.

However, we were unable to obtain information relating to utilisation of the NHS App for gender and ethnicity. The FOI response stated that:

> _Information was also requested relating to the gender and ethnicity of NHS App users; however, this is not collected via NHS Login and so is not currently linked to NHS App login activity. Consequently, we are unable to provide login activity by gender and ethnicity of app users at this current time. We currently are working to expand our understanding of NHS App usage by linking to a wider range of demographic attributes and anticipate that we will have a more complete picture of activity by gender and ethnicity as we progress through winter 2025._


## Age Group Analysis
We compared the rate of unique logins per 1,000 population for each of the age groups. To accurately estimate the population for each age group we used estimates from the [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/bulletins/populationestimatesforenglandandwales/mid2023). This revealed for those accessing the NHS App in May 2025 there is significant variation in rates across different age groups.

```{r demo_age_rates_chart}

demo_age_rates_chart

```

However, a better way to look at this would be to compare utilisation with the App to the utilisation of NHS services. This isn't straight-forward though. How do we decide what an appropriate measure of the utilisation of NHS services is? Do we include all types of care activities? Does an admission equate to a GP appointment? Looking at publicly available data we decided to use outpatient attendances by age group as a proxy for healthcare utilisation. We have used the latest available [annual publication on outpatient attendances](https://digital.nhs.uk/data-and-information/publications/statistical/hospital-outpatient-activity/2023-24), which covers the 2023/24 financial year.

Using this data it is possible to compare the proportion of logins by age group, with the proportion of outpatient appointments. The ratio of these two numbers determines if the number of logins to the App for that age group are proportionate to its share of the outpatient activity. For example, if 11% of all logins were for the 20-24 year old's but they represent 15% of the outpatient appointments, then their ratio will be less than 1. This means this age-group would be underrepresented by those using the NHS App.

The charts below show the ratio of unique logins to outpatient attendances for those aged 15+. The vertical line at 1 is used to help identify age groups that are over or underrepresented.

:::{.callout-note}
As the Outpatient Attendance data is only available in 5-year age bands we are unable to include the _13 to 14_ age group from the NHS App login data.
:::

```{r demo_age_uu_login_ratio_chart}

demo_age_op_ratio_chart

```

This chart clearly demonstrates that patients who are aged 70 years or older are **underrepresented** in using the NHS App when using outpatient attendances as a proxy for healthcare utilisation.

When comparing with ratio of the proportion of _total logins_ rather than _unique logins_ to the proportion of outpatient attendances we see a similar picture:

```{r demo_Age_total_login_ratio_chart}

demo_age_op_logins_ratio_chart

```


## Deprivation Analysis

::: {.callout-warning}
## Deprivation Deciles
As part of the FOI we were also provided with login statistics broken down by IMD decile. However, this breakdown was not based on the decile **the patient was resident in** but on the decile based on the **location of their GP Practice**. 

:::

Looking at the rate of unique logins per 1,000 population across IMD Deciles showed that there is a higher utilisation of the NHS App within the more deprived areas. However, the variation seen across the deciles is much lower than across age groups.

```{r demo_dep_rates_chart}

demo_dep_rates_chart

```

When we compare the ratio of unique logins to outpatient attendances, we see a similar picture of the more deprived areas having a higher than proportionate usage of the NHS App.

```{r demo_dep_uu_login_ratio_chart}

demo_dep_op_ratio_chart

```

Interpreting this information is difficult, in part due to the caveat that deprivation is based on the location of the GP practice for App utilisation. As we are also analysing these two demographics in isolation we can't answer if the lower rates of utilisation in the most affluent areas are being driven by lower rates amongst older patients. We know that the proportion of patients aged 65+ years across England is higher in more affluent areas but without more granular data this can't be fully answered.

# Going forwards
***

This analysis raises some key questions for NHS England. The NHS App has the potential for members of the public to offer more insightful feedback about the care they receive. However, there needs to be the assurance that the NHS is getting feedback from all those who use its services.

## Monitoring App Utilisation
Ensuring that there is a greater understanding of who is using the App is imperative. Whether this is through better data collection via the App or improved data linkage, this will need to happen as a priority. Understanding in more depth if the use of the App is representative of those who use NHS services is the first step to having the assurance feedback won't be biased in favour of certain demographics.

NHS England did state in response to the FOI that: 

> _We currently are working to expand our understanding of NHS App usage by linking to a wider range of demographic attributes and anticipate that we will have a more complete picture of activity by gender and ethnicity as we progress through winter 2025._

Making as much of this data publicly available will also be important to provide the wider public with the confidence that the NHS is working to ensure that this is being monitored and acted on.

## Engaging with different groups
Since their creation in 2012, Healthwatch, both local and national, has gathered patient feedback on how NHS services are performing and understanding what matters to patients. Local Healthwatch organisations have a history of engaging with _seldom heard groups_ to ensure that their views on health and care services aren't ignored.

The 10-year plan also announced the abolition of Healthwatch England and local Healthwatch engagement functions becoming the responsibility of ICBs and providers. There is a risk that feedback on care from these groups via the NHS App isn't aggregated together, and that these groups lose any collective voice. Therefore, ensuring that ICB and providers engage with these groups remains a priority.

## Responsibility for App Engagement
The 10-year plan states that the NHS will:

> _We will also continue our current partnership with libraries and other community organisations to help set people up on the NHS App. We will continue recruiting App Ambassadors across the country to support uptake._

[NHS App Ambassadors](https://digital.nhs.uk/services/nhs-app/become-an-nhs-app-ambassador) are advocates for the NHS App which can include delivering training individuals how to download and use the App. Recruitment of NHS App Ambassadors is currently the responsibility of NHS England. Given the abolition of NHS England, there remains the question of who has responsibility for this going forwards. Will this transition to the Department for Health and Social Care or become the responsibility of local care organisations? Whatever the direction on this, it will be vital that information about who is using the App can be shared with local teams and App Ambassadors. 


# Metadata
***

As part of our commitment to transparency in our work all of the data and code that is used to create this analysis is available from the associated [GitHub repository](https://github.com/NHS-Transformation-Unit/nhs_app_utilisation). This includes specific documentation on data sources saved within the `documentation` folder.
